---
title: 'Linear Regression: A Revisit'
author: "Linh Tran"
date: "1/26/2021"
output: 
   html_document:
     toc: true
---

```{r setup, include=FALSE}

install.packages("summarytools", repos = "http://cran.us.r-project.org")
install.packages("RNHANES", repos = "http://cran.us.r-project.org")

library(RNHANES)
library(tidyverse)
library(summarytools)
library(leaps)
```

# Data

In this example, we assess the association between high density lipoprotein (HDL) cholesterol and BMI, and other demographic factors (age, gender, race) using the NHANES data (https://wwwn.cdc.gov/nhcs/nhanes/ContinuousNhanes/Defaults.aspx?BeginYear=2001). The data can be downloaded using functions in the package RNHANES.

```{r}
dat = nhanes_load_data(file_name = "l13_B", year = "2001-2002")

dat = dat %>% 
  left_join(nhanes_load_data("BMX_B", "2001-2002"), by="SEQN")%>%
  left_join(nhanes_load_data("BPX_B", "2001-2002"), by="SEQN")%>%
  left_join(nhanes_load_data("DEMO_B", "2001-2002"), by="SEQN")

dat = dat%>%
  select(SEQN, RIAGENDR, RIDRETH1, RIDAGEYR, BMXBMI, BPXSY1, LBDHDL)%>%
  mutate(RIAGENDR =as_factor(RIAGENDR), RIDRETH1 =as_factor(RIDRETH1))

colnames(dat) <- c("ID", "gender", "race", "age", "bmi", "sbp", "hdl")

dat <- na.omit(dat)

dat
```

Summary statistics of the predictors and the response

```{r}
st_options(plain.ascii = FALSE,
           style = "rmarkdown",
           dfSummary.silent = TRUE,
           footnote = NA,
           subtitle.emphasis = FALSE)

dfSummary(dat[, -1])
```



# Multiple Linear Regression: A small experiment


```{r}
fit1 <-lm(hdl ~ bmi+age+gender+race+sbp,
          data = dat)
fit1

fit2 <-lm(bmi ~ age+gender+race+sbp,
          data = dat)
fit2

r2 <- fit2$residuals
r2 #residuals from second model
mean(r2)  #close to 0

fit3 <-lm(hdl ~ r2,data = dat)
fit3  #same coefficients as bmi

coef(fit1)["bmi"]  
coef(fit3)["r2"]  

```


If bmi highly correlated to any of the other variables => residual from the fit2 will be very close to 0. => If we regress Y on smt that is very highly correlated (residual close to 0) then variance will be very large.
\
2-step procedures: (1) use bmi as response variable and get residual, (2) simple linear regression with residual as variable and get the same coefficient. => impact of correlation  

# Prediction interval vs. Confidence interval


```{r}
newdata = dat[1, ] #just looking at 1st observation

predict(fit1, newdata, interval = "confidence")


predict(fit1, newdata, interval = "predict") #much wider


```


# Best subset model selection

```{r}
regsubsets0bj = regsubsets(hdl ~ bmi + age + gender + race + sbp, data = dat,
                           method = "exhaustive", nbest = 1)

plot(regsubsets0bj, scale = "bic")
```

We want BIC as high as possible. This just provides a reference but shouldn't based on this to select the final model. 
