---
title: "Ridge Regression and Lasso"
author: "Yifei Sun"
output:
  pdf_document:
    toc: yes
    toc_depth: 2
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
header-includes:
- \usepackage{fancyhdr}
- \usepackage{lipsum}
- \pagestyle{fancy}
- \fancyhead[R]{\thepage}
- \fancypagestyle{plain}{\pagestyle{fancy}}
--- 

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(ISLR)
library(glmnet)
library(caret)
library(corrplot)
library(plotmo)
```

Predict a baseball player’s salary on the basis of various statistics associated with performance in the previous year. Use `?Hitters` for more details.

```{r}
#Load data from ISLR package
data(Hitters)  

# delete rows containing the missing data
Hitters <- na.omit(Hitters)
Hitters2 <- model.matrix(Salary ~ ., Hitters)[ ,-1]  #remove 1st column

set.seed(1)
trainRows <- createDataPartition(y = Hitters$Salary, p = 0.8, list = FALSE)

# matrix of predictors (glmnet uses input matrix)
x <- Hitters2[trainRows,]
# vector of response
y <- Hitters$Salary[trainRows]

corrplot(cor(x), method = "circle", type = "full")

corrplot(cor(x), method = "circle", type = "upper")
#when visualize general matrix, is.corr = FALSE
```

# Using `glmnet`
## Ridge 

`alpha` is the elastic net mixing parameter. `alpha=1` is the lasso penalty, and `alpha=0` the ridge penalty. `glmnet()` function standardizes the independent variables by default (The coefficients are always returned on the original scale). 

```{r}
# fit the ridge regression (alpha = 0) with a sequence of lambdas
ridge.mod <- glmnet(x = x, y = y, 
                    standardize = TRUE,
                    alpha = 0, 
                    lambda = exp(seq(10, -2, length = 100)))
```

`coef(ridge.mod)` gives the coefficient matrix. Each column is the fit corresponding to one lambda value.

```{r}
mat.coef <- coef(ridge.mod)
dim(mat.coef)
```


### Trace plot
There are two functions for generating the trace plot.
```{r}
# plot(ridge.mod, xvar = "lambda", label = TRUE)
plot_glmnet(ridge.mod, xvar = "rlambda", label = 19)
```

### Cross-validation
We use cross-validation to determine the optimal value of `lambda`. The two vertical lines are the for minimal MSE and 1SE rule. The 1SE rule gives the most regularized model such that error is within one standard error of the minimum.

```{r}
set.seed(2)
cv.ridge <- cv.glmnet(x, y, 
                      # type.measure = "mse",
                      alpha = 0, 
                      lambda = exp(seq(10, -2, length = 100)))

# set.seed(2)
# cv.ridge <- cv.glmnet(x, y, alpha = 0, nlambda = 200)

plot(cv.ridge)
abline(h = (cv.ridge$cvm + cv.ridge$cvsd)[which.min(cv.ridge$cvm)], col = 4, lwd = 2)

# min CV MSE
cv.ridge$lambda.min

# the 1SE rule
cv.ridge$lambda.1se
```


### Coefficients of the final model

Get the coefficients of the optimal model. `s` is value of the penalty parameter `lambda` at which predictions are required.

```{r}
# extract coefficients
predict(cv.ridge, s = cv.ridge$lambda.min, type = "coefficients") 

# make prediction
head(predict(cv.ridge, newx = Hitters2[-trainRows,], 
             s = "lambda.min", type = "response")) 

# predict(cv.ridge, s = "lambda.min", type = "coefficients") 
# predict(cv.ridge, s = "lambda.1se", type = "coefficients") 
# predict(ridge.mod, s = cv.ridge$lambda.min, type = "coefficients")
```


## Lasso 
The syntax is along the same line as ridge regression. Now we use `alpha = 1`.

```{r}
set.seed(2)  #if compare the models, MUST set seed

cv.lasso <- cv.glmnet(x, y, 
                      alpha = 1, 
                      lambda = exp(seq(5, -1, length = 100)))

cv.lasso$lambda.min
```

```{r}
#CV plot 
plot(cv.lasso)
```

```{r}
# cv.lasso$glmnet.fit is a fitted glmnet object using the full training data
# plot(cv.lasso$glmnet.fit, xvar = "lambda", label=TRUE)
plot_glmnet(cv.lasso$glmnet.fit)
```


```{r}
predict(cv.lasso, s = "lambda.min", type = "coefficients")

head(predict(cv.lasso, newx = Hitters2[-trainRows,], s = "lambda.min", type = "response"))
```

# Using `caret` 

## Ridge

```{r}
ctrl1 <- trainControl(method = "repeatedcv", number = 10, repeats = 5) #repeat 10fold CV 5 times

# you can try other options

set.seed(2)
ridge.fit <- train(x, y,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 0, 
                                          lambda = exp(seq(10, -2, length=100))),
                   # preProc = c("center", "scale"),
                   trControl = ctrl1)

plot(ridge.fit, xTrans = log)

ridge.fit$bestTune

# extract coefficients in the final model
coef(ridge.fit$finalModel, s = ridge.fit$bestTune$lambda)
```

## Lasso

```{r}
set.seed(2)
lasso.fit <- train(x, y,
                   method = "glmnet",
                   tuneGrid = expand.grid(alpha = 1, 
                                          lambda = exp(seq(5, -1, length=100))),
                   trControl = ctrl1)
plot(lasso.fit, xTrans = log)

lasso.fit$bestTune

coef(lasso.fit$finalModel, lasso.fit$bestTune$lambda)
```

## Elastic net

For elastic net, lambda is somewhere between 0 and 1 => need to use expand.grid

```{r}
set.seed(2)
enet.fit <- train(x, y,
                  method = "glmnet",
                  tuneGrid = expand.grid(alpha = seq(0, 1, length = 11), 
                                         lambda = exp(seq(2, -2, length = 50))),
                  trControl = ctrl1)
enet.fit$bestTune

plot(enet.fit)

coef(enet.fit$finalModel, enet.fit$bestTune$lambda)
```

## Comparing different models

```{r, fig.width=5}
set.seed(2)
lm.fit <- train(x, y,
                method = "lm",
                trControl = ctrl1)

resamp <- resamples(list(enet = enet.fit, lasso = lasso.fit, ridge = ridge.fit, lm = lm.fit)) #enet give the smallest MSE
summary(resamp)

bwplot(resamp, metric = "RMSE")
```

## Prediction

```{r}
enet.pred <- predict(enet.fit, newdata = Hitters2[-trainRows,])
# test error
mean((enet.pred - Hitters$Salary[-trainRows])^2)

sqrt(mean((enet.pred - Hitters$Salary[-trainRows])^2))
```