---
title: "Linear Methods for Classification I"
author: "Yifei Sun"
output: html_document

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo = T, message = FALSE, results='hide', warning=FALSE}
library(caret)
library(glmnet)
library(mlbench)  #diabetes dataset
library(pROC)   #used to generate ROC curve, calculate area under the curve
library(pdp) 
library(vip)  #create very important plots
library(AppliedPredictiveModeling)
library(e1071)
```

We use the Pima Indians Diabetes Database for illustration. The data contain 768 observations and 9 variables. The outcome is a binary variable `diabetes`. We start from some simple visualization of the data.

```{r}
data(PimaIndiansDiabetes)
dat <- PimaIndiansDiabetes
head(dat)

theme1 <- transparentTheme(trans = .4)  #transparent color
trellis.par.set(theme1)

featurePlot(x = dat[, 1:8], 
            y = dat$diabetes,
            scales = list(x = list(relation = "free"), 
                          y = list(relation = "free")),
            plot = "density", pch = "|", 
            auto.key = list(columns = 2)
            )

   
summary(dat) 

```


The data is divided into two parts (training and test). 
```{r}
set.seed(1)

rowTrain <- createDataPartition(y = dat$diabetes,
                                p = 0.75,
                                list = FALSE)
```

# Logistic regression and its cousins

## `glm`

```{r}
contrasts(dat$diabetes)

glm.fit <- glm(diabetes ~ ., 
               data = dat, 
               subset = rowTrain, 
               family = binomial(link = "logit"))
               
summary(glm.fit)
```



We first consider the simple classifier with a cut-off of 0.5 and evaluate its performance on the test data.

```{r}
test.pred.prob <- predict(glm.fit, newdata = dat[-rowTrain,],
                           type = "response")
                           
head(test.pred.prob)

test.pred <- rep("neg", length(test.pred.prob))
test.pred[test.pred.prob>0.5] <- "pos"
head(test.pred)

# 2x2 table
confusionMatrix(data = as.factor(test.pred),
                reference = dat$diabetes[-rowTrain],
                positive = "pos")
```

We then plot the test ROC curve. You may also consider a smoothed ROC curve.

```{r}
roc.glm <- roc(dat$diabetes[-rowTrain], test.pred.prob)
plot(roc.glm, legacy.axes = TRUE, print.auc = TRUE)
plot(smooth(roc.glm), col = 4, add = TRUE)
```

We can also fit a logistic regression using caret. This is to compare the cross-validation performance with other models, rather than tuning the model.

```{r}
# Using caret
ctrl <- trainControl(method = "cv",
                     summaryFunction = twoClassSummary, #don't use if have multiple classes
                     classProbs = TRUE)
set.seed(1)
model.glm <- train(x = dat[rowTrain,1:8],
                   y = dat$diabetes[rowTrain],
                   method = "glm",
                   metric = "ROC",
                   trControl = ctrl)
```

## Penalized logistic regression

Penalized logistic regression can be fitted using `glmnet`. We use the `train` function to select the optimal tuning parameters.

```{r}
glmnGrid <- expand.grid(.alpha = seq(0, 1, length = 6),
                        .lambda = exp(seq(-8, -2, length = 20)))
set.seed(1)
model.glmn <- train(x = dat[rowTrain,1:8],
                    y = dat$diabetes[rowTrain],
                    method = "glmnet",
                    tuneGrid = glmnGrid,
                    metric = "ROC",
                    trControl = ctrl)

plot(model.glmn, xTrans = function(x) log(x))   

model.glmn$bestTune
```

## GAM

```{r}
set.seed(1)
model.gam <- train(x = dat[rowTrain,1:8],
                   y = dat$diabetes[rowTrain],
                   method = "gam",
                   metric = "ROC",
                   trControl = ctrl)


model.gam$finalModel

# plot(model.gam$finalModel)
```


## MARS

```{r}
set.seed(1)
model.mars <- train(x = dat[rowTrain,1:8],
                    y = dat$diabetes[rowTrain],
                    method = "earth",
                    tuneGrid = expand.grid(degree = 1:3, 
                                           nprune = 2:15),
                    metric = "ROC",
                    trControl = ctrl)

plot(model.mars)

coef(model.mars$finalModel) 

pdp::partial(model.mars, pred.var = c("age"), grid.resolution = 200) %>% autoplot()

vip(model.mars$finalModel)
```


```{r, warning=FALSE}
set.seed(1)

model.knn <- train(x = dat[rowTrain,1:8],
                   y = dat$diabetes[rowTrain],
                   method = "knn",
                   preProcess = c("center","scale"),
                   tuneGrid = data.frame(k = seq(1,200,by=5)),
                   trControl = ctrl)

ggplot(model.knn, highlight = TRUE)
```


```{r}
res <- resamples(list(GLM = model.glm, 
                      GLMNET = model.glmn, 
                      GAM = model.gam,
                      MARS = model.mars,
                      KNN = model.knn))
summary(res)

bwplot(res, metric = "ROC")
```

Now let's look at the test data performance.
```{r, warning=FALSE}
glm.pred <- predict(model.glm, newdata = dat[-rowTrain,], type = "prob")[,2]
glmn.pred <- predict(model.glmn, newdata = dat[-rowTrain,], type = "prob")[,2]
knn.pred <- predict(model.knn, newdata = dat[-rowTrain,], type = "prob")[,2]
gam.pred <- predict(model.gam, newdata = dat[-rowTrain,], type = "prob")[,2]
mars.pred <- predict(model.mars, newdata = dat[-rowTrain,], type = "prob")[,2]

roc.glm <- roc(dat$diabetes[-rowTrain], glm.pred)
roc.glmn <- roc(dat$diabetes[-rowTrain], glmn.pred)
roc.knn <- roc(dat$diabetes[-rowTrain], knn.pred)
roc.gam <- roc(dat$diabetes[-rowTrain], gam.pred)
roc.mars <- roc(dat$diabetes[-rowTrain], mars.pred)

auc <- c(roc.glm$auc[1], roc.glmn$auc[1], roc.knn$auc[1],
         roc.gam$auc[1], roc.mars$auc[1])

plot(roc.glm, legacy.axes = TRUE)
plot(roc.glmn, col = 2, add = TRUE)
plot(roc.knn, col = 3, add = TRUE)
plot(roc.gam, col = 4, add = TRUE)
plot(roc.mars, col = 5, add = TRUE)

modelNames <- c("glm","glmn","knn","gam","mars")
legend("bottomright", legend = paste0(modelNames, ": ", round(auc,3)),
       col = 1:5, lwd = 2)
```



```{r}
summary(Smarket)
length(Smarket$Direction) #1250

glm.fit=glm(Direction∼Lag1+Lag2+Lag3+Lag4+Lag5+Volume ,
data=Smarket ,family=binomial )

summary (glm.fit)

glm.probs= predict(glm.fit ,type="response")
length(glm.probs)  #3686

glm.pred=rep("Down" ,1250)
length(glm.pred)  #1250

glm.pred[glm.probs >.5]=" Up"

table(glm.pred, Smarket$Direction)

```


